Prestige bias in cultural evolutionary dynamics

### 2023-11-08 ms/ms.tex
finished with rev1 revisions

Boyd and Richerson~\citep[Ch.~5]{evolutionBook} suggested that the assessment of success can be divided to three categories: \emph{direct bias}, \emph{indirect bias} and \emph{frequency-dependent bias}.
Direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \emph{directly} testing the trait.
For example, an individual observing a ping-pong match can try the observed paddle grips to determine which grip is better.
Frequency-dependent bias occurs when the probability of copying a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, suppose the common paddle grip is used by 60\% of the demonstrators; if the this grip is adopted by 80\% of copiers, then transmission is under positive frequency bias, also called \emph{conformity}; if it is adopted by 40\% of copiers, then transmission is under negative frequency bias, or \emph{anti-conformity} \citep{conformism}. 
The effects of conformity and anti-conformity on cultural evolution have been studied with both models \citep{anticonformity,Denton2021,conformity} and experiments \citep{negativeFrequency}.
Indirect bias occurs when a copier uses some observed phenotype to evaluate the attractiveness of a potential role-model. 
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
However, this may cause mismatches between the copied trait and the rest of the cultural or genetic repertoire of the individual \citep{Kolodny2022}.
Furthermore, Boyd and Richerson~\citep[Ch. 8]{evolutionBook} suggest that maladaptive traits may spread widely in a population if indirect bias is strong enough, e.g., by a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
Indeed, helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient~\citep{cooperation}.

@@

The distinction between success and prestige bias is important, as prestige is a context-dependent bias, rather than a content-dependent bias: it does not depend on the phenotype itself but rather on the number of copiers that have already copied each role-model, which may be easier and more accurate to estimate than success. 
Prestige bias is also frequency independent (see \Cref{cor:GBD} below), and thus it differs from conformity \citep{anticonformity,Denton2021,conformity}, which depends on the frequency of a trait in the population or in a sample of role-models, rather than the social dynamics of copying.

@@

\subsection*{Continuous trait}
We follow the Boyd and Richerson model~\citep{evolutionBook}, assuming only oblique transmission of a single trait. 
Consider a population of $N$ individuals, described by a single trait that takes continuous values.
Each generation, $N$ naive individuals, or copiers, each chooses a single role-model from the entire previous generation. The copier then copies its trait value from the chosen role-model.
Note that our transmission models are slightly different from those modeled before, e.g. \citep{cumul_culture,anticonformity,evolutionBook}, in which the population is infinite and each copier samples $n$ role-models and then copies its trait from one or more of the sampled role-models.

@@

Hence, our model has two ``nested'' stochastic processes: the role-model choice process within each generation, and the cultural-evolutionary process between generations.
To simplify the mathematical and computational analysis, we developed analytic approximations for the role-model choice process using the the {\em generalized binomial distribution} (GBD, \Cref{res:GBD}) and the {\em Dirichlet-Multinomial distribution} (DM, \Cref{cor:dirichlet}).
The latter is especially useful, as it approximates the entire role-model choice process and only requires us to assume that the relative effect of success and influence is a characteristic of the role-model and not the copier.

Analyzing the model with the DM distribution, we found approximations for the fixation probability and fixation time of a cultural trait under biased transmission in a constant environment.
Our approximations are similar to Kimura's evolutionary-genetic approximations, in that (i) the difference between the resident and invading cultural trait values, $1-\beta(A)$, is equivalent to the selection coefficient in favor of a beneficial allele, $s$, and (ii) increasing the relative weight of influence versus success bias, $\alpha$, decreases the effective population size, $N_e$~(\Cref{fig:var_alpha}).

@@

Lastly, we examined a scenario in which copiers can adjust their success-weight bias, $\alpha$, to minimize their copying error, i.e., copy trait values closer to the optimal value.
We found that as the role-model choice process proceeds (that is, more copiers make their choices), both the success-bias weight (adjusted by copiers) and the estimation error decrease. 
The latter is significantly lower than in a population using a constant, fixed success-bias weight, regardless of the value of the constant weight (\Cref{fig:influence_advantage}).
This suggests that the later a copier makes its choice, the more it should rely on choices of previous copiers (prestige), and the less it should rely on its own estimation of the success of role-models.
The rationale, then, is that the more information a copier has, e.g., by using others as information sources, the more informative and effective his choice can be.

Others have analyzed models with interactions between different transmission biases.
Ze Hong~\citep{payoff} studied a model with both frequency-dependent bias (including conformist bias) and payoff bias. The analysis demonstrates that an integrated learning strategy may be superior to relying on either bias alone. Moreover, both biases are likely to contribute to the spread of variants with high payoff, and the magnitude of conformist bias affects the evolutionary dynamics.
% TODO add: 
% Ammar, Madeleine, Laurel Fogarty, and Anne Kandler. 2023. “Social Learning and Memory.” Proceedings of the National Academy of Sciences 120 (33): e2310033120. https://doi.org/10.1073/pnas.2310033120.
% Enquist, Magnus, and Stefano Ghirlanda. 2007. “Evolution of Social Learning Does Not Explain the Origin of Human Cumulative Culture.” Journal of Theoretical Biology 246 (1): 129–35. https://doi.org/10.1016/j.jtbi.2006.12.022.
% Denton, Kaleda K., Yoav Ram, and Marcus W. Feldman. 2022. “Conformity and Content-Biased Cultural Transmission in the Evolution of Altruism.” Theoretical Population Biology 143 (February): 52–61. https://doi.org/10.1016/j.tpb.2021.10.004.

@@

According to Henrich and Broesch~\citep{fijian_social_bias}, natural selection has favored the emergence of psychological biases for learning from those individuals most likely to possess adaptive information. 
They authors studied Fijian villages to examine if and how such biases emerge in a small-scale society.
They found that Fijian villagers are more likely to learn from role-models perceived as more successful/knowledgeable, both within and across domains. Their research thus suggests that copying from those perceived as successful, rather than who are actually successful, is a common phenomenon. They show that the social networks representing copier--role-model relationships are centralized, which is consistent with the prediction that people substantially share notions about who is a good cultural model, but that individuals' role-model selections are influenced by multiple factors.

@@

@@

\paragraph{Conclusions.}
We studied a model of cultural evolution under two transmission biases: the commonly studied success bias, together with prestige bias, which has so far received less attention. We found approximations for this complex dynamics. We then showed that success bias affects the evolutionary dynamics much like natural selection does, whereas prestige bias has a similar effect to random genetic drift. We also find a clear advantage to individuals that can choose the relative weight of the two biases.

{\small
\section*{Acknowledgements}
We thank Marc Feldman, Martin Pontz, Gili Greenbaum, Alon Rozen, Tal Simon and two anonymous reviewers for discussions and comments.
This work was supported in part by 
Minerva Stiftung Center for Lab Evolution (YR) and  the John Templeton Foundation (YR 61809).
}

@@

\bibitem{conformity}
Denton, Kaleda K., Yoav Ram, and Marcus W. Feldman. (2022) "Conformity and content-biased cultural transmission in the evolution of altruism." Theoretical Population Biology 143: 52-61.

\bibitem{anticonformity}
Denton, Kaleda Krebs, Yoav Ram, Uri Liberman, and Marcus W. Feldman. (2020) "Cultural Evolution of Conformity and Anticonformity." Proceedings of the National Academy of Sciences 117 (24): 13603–14.

@@

\bibitem{Fogarty2017}Fogarty, L., J. Y. Wakano, M. W. Feldman, and K. Aoki. (2017) The driving forces of cultural complexity: Neanderthals, modern humans, and the question of population size. Human Nature : 28: 39-52.

\bibitem{cumul_culture}Denton, Kaleda K, Yoav Ram, and Marcus W. Feldman. (2023) "Conditions That Favour Cumulative Cultural Evolution." Philosophical Transactions of the Royal Society B: Biological Sciences 378 (1872).

@@

>>>>

Boyd and Richerson~\citep[Ch.~5]{evolutionBook} suggested that the assessment of success can be divided to three categories: \emph{direct bias}, \emph{indirect bias} and \emph{frequency-dependent bias}.
Direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \emph{directly} testing the trait.
For example, an individual observing a ping-pong match can try the observed paddle grips to determine which grip is better.
Frequency-dependent bias occurs when the probability of copying a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, suppose the common paddle grip is used by 60\% of the demonstrators; if the this grip is adopted by 80\% of copiers, then transmission is under positive frequency bias, also called \emph{conformity}; if it is adopted by 40\% of copiers, then transmission is under negative frequency bias, or \emph{anti-conformity} \citep{conformism}. 
The effects of conformity and anti-conformity on cultural evolution have been studied with both models \citep{Denton2020,Denton2021,Denton2022} and experiments \citep{negativeFrequency}.
Indirect bias occurs when a copier uses some observed phenotype to evaluate the attractiveness of a potential role-model. 
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
However, this may cause mismatches between the copied trait and the rest of the cultural or genetic repertoire of the individual \citep{Kolodny2022}.
Furthermore, Boyd and Richerson~\citep[Ch. 8]{evolutionBook} suggest that maladaptive traits may spread widely in a population if indirect bias is strong enough, e.g., by a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
Indeed, helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient~\citep{cooperation}.

@@

The distinction between success and prestige bias is important, as prestige is a context-dependent bias, rather than a content-dependent bias: it does not depend on the phenotype itself but rather on the number of copiers that have already copied each role-model, which may be easier and more accurate to estimate than success. 
Prestige bias is also frequency independent (see \Cref{cor:GBD} below), and thus it differs from conformity \citep{Denton2020,Denton2021,Denton2022}, which depends on the frequency of a trait in the population or in a sample of role-models, rather than the social dynamics of copying.

@@

\subsection*{Continuous trait}
We follow the Boyd and Richerson model~\citep{evolutionBook}, assuming only oblique transmission of a single trait. 
Consider a population of $N$ individuals, described by a single trait that takes continuous values.
Each generation, $N$ naive individuals, or copiers, each chooses a single role-model from the entire previous generation. The copier then copies its trait value from the chosen role-model.
Note that our transmission models are slightly different from those modeled before, e.g. \citep{cumul_culture,Denton2021,evolutionBook}, in which the population is infinite and each copier samples $n$ role-models and then copies its trait from one or more of the sampled role-models.

@@

Hence, our model has two ``nested'' stochastic processes: the role-model choice process within each generation, and the cultural-evolutionary process between generations.
To simplify the mathematical and computational analysis, we developed analytic approximations for the role-model choice process using the the {\em generalized binomial distribution} (GBD, \Cref{res:GBD}) and the {\em Dirichlet-Multinomial distribution} (DM, \Cref{cor:dirichlet}).
The latter is especially useful, as it approximates the entire role-model choice process and only requires us to assume that the relative effect of success and prestige is a characteristic of the role-model and not the copier.

Analyzing the model with the DM distribution, we found approximations for the fixation probability and fixation time of a cultural trait under biased transmission in a constant environment.
Our approximations are similar to Kimura's evolutionary-genetic approximations, in that (i) the strength of success bias towards the invading cultural trait, $1-\beta(A)$, is equivalent to the selection coefficient in favor of a beneficial allele, $s$, and (ii) decreasing the relative weight of success versus prestige bias, $\alpha$, decreases the effective population size, $N_e$~(\Cref{fig:var_alpha}).

@@

Lastly, we examined a scenario in which copiers can adjust their success-bias weight, $\alpha$, to minimize their copying error, i.e., copy trait values closer to the optimal value.
We found that as the role-model choice process proceeds (that is, more copiers make their choices), both the success-bias weight (adjusted by copiers) and the estimation error decrease. 
The latter is significantly lower than in a population using a constant, fixed success-bias weight, regardless of the value of the constant weight (\Cref{fig:influence_advantage}).
This suggests that the later a copier makes its choice, the more it should rely on choices of previous copiers (prestige), and the less it should rely on its own estimation of the success of role-models.
The rationale, then, is that the more information a copier has, e.g., by using others as information sources, the more informative and effective his choice can be.

@@

According to Henrich and Broesch~\citep{fijian_social_bias}, natural selection has favored the emergence of psychological biases for learning from those individuals most likely to possess adaptive information. 
The authors studied Fijian villages to examine if and how such biases emerge in a small-scale society.
They found that Fijian villagers are more likely to learn from role-models perceived as more successful/knowledgeable, both within and across domains. Their research thus suggests that copying from those perceived as successful, rather than who are actually successful, is a common phenomenon. They show that the social networks representing copier--role-model relationships are centralized, which is consistent with the prediction that people substantially share notions about who is a good cultural model, but that individuals' role-model selections are influenced by multiple factors.

@@

Others have analyzed models with interactions between different transmission biases.
Hong~\citep{payoff} studied a model with both conformity and success bias (which he calls `payoff bias'). 
He showed that an intermediate level of conformity bias--not too little but not too much--can be adaptive and evolve to prevent invasion of low-success traits while allowing the invasion of high-success traits (for another example of adaptive filtering, see~\citep{adaptive_filtering}).
Similar to our model (\cref{eq:prestige}), Hong~\citep{payoff} also additively combined the two transmission biases (his eq.~1). However, transmission biases can be combined in many ways. 
For example, Denton et al.~\citep{Denton2022} combined frequency-dependent bias and genetically determined content bias multiplicatively (their eq.~1). 
Ammar et al.~\citep{Ammar2023} studied a model in which individuals have a repertoire of cultural variants to choose from, and both variant choice and transmission via social learning are success-biased. Moreover, they also included the possibility to `forget' infrequently used variants; therefore, because usage is success-biased, memory is also success-biased. It remains to be seen how different assumptions on the mechanisms of learning and forgetting affect the evolutionary dynamics under different and interacting transmission biases.

@@

\paragraph{Conclusions.}
We studied a model of cultural evolution under two transmission biases: the commonly studied success bias, together with prestige bias, which has so far received less attention by modelers. We found approximations for this complex dynamics. We then showed that success bias affects the evolutionary dynamics much like natural selection does, whereas prestige bias has a similar effect to random genetic drift. We also find a clear advantage to individuals that can choose the relative weight of the two biases.

{\small
\section*{Acknowledgements}
We thank Marc Feldman, Martin Pontz, Gili Greenbaum, Alon Rozen, Tal Simon and two anonymous reviewers for discussions and comments.
This work was supported in part by 
Minerva Stiftung Center for Lab Evolution~(YR) and  the John Templeton Foundation~(YR 61809).
}

\newpage

@@

\bibitem{Denton2022}
Denton, Kaleda K., Yoav Ram, and Marcus W. Feldman. (2022) "Conformity and content-biased cultural transmission in the evolution of altruism." Theoretical Population Biology 143: 52-61.

\bibitem{Denton2020}
Denton, Kaleda Krebs, Yoav Ram, Uri Liberman, and Marcus W. Feldman. (2020) "Cultural evolution of conformity and anticonformity." Proceedings of the National Academy of Sciences 117 (24): 13603–14.

@@

\bibitem{Fogarty2017} 
Fogarty, L., J. Y. Wakano, M. W. Feldman, and K. Aoki. (2017) The driving forces of cultural complexity: Neanderthals, modern humans, and the question of population size. Human Nature : 28: 39-52.

\bibitem{cumul_culture}
Denton, Kaleda K, Yoav Ram, and Marcus W. Feldman. (2023) "Conditions That Favour Cumulative Cultural Evolution." Philosophical Transactions of the Royal Society B: Biological Sciences 378 (1872).

@@

\bibitem{adaptive_filtering}
 Enquist, Magnus, and Stefano Ghirlanda. (2007) "Evolution of Social Learning Does Not Explain the Origin of Human Cumulative Culture." Journal of Theoretical Biology 246 (1): 129–35.

\bibitem{Ammar2023}
Ammar, Madeleine, Laurel Fogarty, and Anne Kandler. (2023) "Social Learning and Memory." Proceedings of the National Academy of Sciences 120 (33): e2310033120.
